Peptide retention time prediction

### 2015-10-12 retention.tex
Submitted version

\documentclass[a4paper]{article}
\usepackage[margin=1in]{geometry} 
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{multirow}

@@

In this review we summarize the work done in the field of retention
time prediction, and describe a few of the successful applications of
such predictions. Section~\ref{sec:rplc} describes the basics of
peptide separation using reversed-phase liquid chromatography, while
Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} summarize the main
approaches to retention time predictions from peptide
sequences. Section~\ref{sec:app} discusses some of the most important
applications of retention time predictions.
% Section~\ref{sec:irt} briefly introduces the use of peptide retention
% standards.

@@

\begin{figure}[!h]
\centering
\includegraphics[trim=0.5cm 2.8cm 0.5cm 3.5cm, clip=true, width=0.8\textwidth]{img/rplc.pdf}
\caption{\label{fig:rplc} {\bf Principles of liquid chromatography.}
  The peptides mixed with mobile phase are injected on the column
  (Panel A). As an increasing amount of solvent B is mixed in, the
  peptides are released from the column in order of hydrophobicity
  (Panels B and C). The signal at the end of the column can be
  measured at each time point and displayed as a chromatogram (Panel
  D). Image adapted from www.lamondlab.com.}
\vspace{-7pt}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
\caption{\label{fig:repr} {\bf Reproducibility of peptide retention
    time.} Two identical shotgun proteomics experiments were carried
  out. The retention times of the peptides confidently identified in
  both runs are displayed. It should be noted that this relatively high reproducibility normally only appears between runs under near to identical experiments, and do not carry over into other experimental conditions or instrumentation.}
\end{figure}

@@

\begin{figure}[!h]
\centering \includegraphics[clip=true,
  width=0.8\textwidth]{img/methods.pdf}
\caption{\label{fig:methods} {\bf Retention time prediction methods.}
  Each figure illustrates a class of retention time prediction
  methods. The details of the different methods are described in
  detail in Section \ref{sec:rtpred}}.
\vspace{-7pt}
\end{figure}

@@

\begin{table}
 \caption{{\bf Retention time prediction methods.} We
   give a few of the most commonly used retention time prediction methods available.}
 \vspace{.2cm}
 \label{tab:rtmethods}
 \begin{tabular}{lp{9cm}l}

@@

A limitation of index-based methods are that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
Elude} \cite{elude1,elude2} estimates retention time indices for each
dataset as hand. In these cases the indeces are used to generate
features that are used as an input for final retention time
estimates. We will come back to this in section \ref{sec:ML}.

@@

In a follow-up work, the method was extended by representing each
peptide by 1052 instead of 20 features
\cite{petritis2006improved}. The new descriptors encoded the identity
of each amino acid at each position in a sequence, as well as the
length and the hydrophobic moment of the peptide. The new neural
network was trained on more than 300,000 peptide sequences. To obtain
such a large number of peptides with known retention times, a strategy
to normalize elution times was developed. This allowed pooling of
peptides identified in more than 12,000 shotgun experiments. The
trained algorithm reported better performance for the 1303 peptides
on which it was evaluated, than its predecesor.

@@

\begin{figure}[!h]
\centering 
\includegraphics[width=0.9\textwidth]{img/elude-ptm.pdf}
\caption{\label{fig:elude} {\bf {\sc Elude}'s workflow.}  {\sc Elude} encompasses two learning steps. In the first one the training peptides are used to build an in-house retention index. This index, together with additional information, is then employed to calculate a set of 60 numerical features for each peptide. These features are optimally combined during a second training procedure. The resulting retention model can be subsequently used to predict the retention time of other peptides of interest.}
\end{figure}

@@

The available technologies to infer peptides and proteins from mass
spectrometry data involve predicting the the potential manifestations
of the analyzed peptides, such as their fragmentation pattern, and
comparing those predicted properties to the observed data. Such
comparisons are regularly done for discovery proteomics by search
engines. However, an attractive way to increase the sensitivity of the
inference procedure is to also predict properties such as their
iso-electric point~\cite{perez2011silico,branca2014hirief} or of
particular interest for this review, chromatographic retention time
\cite{cerqueira2010mude}. One way to determie a predicatble sparation
techniques usefulness is to estimate its ability to remove candidate
peptides of any detected feature, normally measured by a false
positive ratio (FPR), also known as the methods fall-out. Current
retention time prediction algorithms have an FPR of about
$\frac{1}{5}$, meaning that it can remove about four out of five
candidate peptides for a given feature. This could be compared to the
FPR of about $\frac{1}{200}$ for predictions of high-resolution
iso-electric focusing experiments~\cite{branca2014hirief}. The
retention time prediction have toimprove quite a bit to come to level
with the fall-out of iso-electric focusing, however, almost all mass
spectrometry-based proteomics use reversed phase chromatography, while
very few studies are using iso-electric focusing.

@@

%%  LocalWords:  hydrophobicity chromatography elute peptides physico
%%  LocalWords:  electrospray spectrometry analytical proteomics RPLC
%%  LocalWords:  posttranslational phosphorylation proteome analytes
%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls et
%%  LocalWords:  workflows trypsin hydrophobicities chromatographic
%%  LocalWords:  alkyl peptide's elution SSRCalc BioLCCC RTModel iRT
%%  LocalWords:  RTPredict Krokhin al QSRR Klammer SVMs SVM Moruz pre
%%  LocalWords:  dataset workflow outliers posttranslationally silico
%%  LocalWords:  phosphorylated artifactual phosphopeptides COmbined
%%  LocalWords:  FRActional DIagonal COFRADIC datasets Palmblad
%%  LocalWords:  Strittmatter unfragmented

>>>>

\documentclass[a4paper]{article}
\usepackage[margin=1in]{geometry} 
\usepackage{graphicx}
\usepackage{wrapfig}
\usepackage{url}
\usepackage{multirow}
%\usepackage[nomarkers,figuresonly]{endfloat}
\usepackage[nomarkers,nolists,noheads]{endfloat}

@@

In this review we summarize the work done in the field of retention
time prediction, and describe a few of the successful applications of
such predictions. Section~\ref{sec:rplc} describes the basics of
peptide separation using reversed-phase liquid chromatography, while
Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} summarize the main
approaches to retention time predictions from peptide
sequences. Section~\ref{sec:app} discusses some of the most important
applications of retention time predictions.

% Section~\ref{sec:irt} briefly introduces the use of peptide retention
% standards.

@@

\begin{figure}[p]
\centering
\includegraphics[trim=0.5cm 2.8cm 0.5cm 3.5cm, clip=true, width=0.8\textwidth]{img/rplc.pdf}
\caption{\label{fig:rplc} {\bf Principles of liquid chromatography.}
  The peptides mixed with mobile phase are injected on the column
  (Panel A). As an increasing amount of solvent B is mixed in, the
  peptides are released from the column in order of hydrophobicity
  (Panels B and C). The signal at the end of the column can be
  measured at each time point and displayed as a chromatogram (Panel
  D). Image adapted from www.lamondlab.com.}
\vspace{-7pt}
\end{figure}

@@

\begin{figure}[p]
\centering
\includegraphics[trim=0.5cm 0cm 2cm 1.5cm, clip=true, width=0.43\textwidth]{img/reproducibility.pdf}
\caption{\label{fig:repr} {\bf Reproducibility of peptide retention
    time.} Two identical shotgun proteomics experiments were carried
  out. The retention times of the peptides confidently identified in
  both runs are displayed. It should be noted that this relatively high reproducibility normally only appears between runs under near to identical experiments, and do not carry over into other experimental conditions or instrumentation.}
\end{figure}

@@

\begin{figure}[p]
\centering \includegraphics[clip=true,
  width=0.8\textwidth]{img/methods.pdf}
\caption{\label{fig:methods} {\bf Retention time prediction methods.}
  Each figure illustrates a class of retention time prediction
  methods. The details of the different methods are described in
  detail in Section \ref{sec:rtpred}}.
\vspace{-7pt}
\end{figure}

@@

\begin{table}[p]
 \caption{{\bf Retention time prediction methods.} We
   give a few of the most commonly used retention time prediction methods available.}
 \vspace{.2cm}
 \label{tab:rtmethods}
 \begin{tabular}{lp{9cm}l}

@@

A limitation of index-based methods are that they are optimized for
predicting retention times for certain chromatography setups. Thus,
their prediction ability deteriorates for experimental conditions
diverging from the setup on which the algorithm was optimized
\cite{Spicer2007}. Consequently, more advanced methods such as {\sc
Elude} \cite{elude1,elude2} estimates retention time indices for each
dataset as hand. In these cases the indices are used to generate
features that are used as an input for final retention time
estimates. We will come back to this in section \ref{sec:ML}.

@@

In a follow-up work, the method was extended by representing each
peptide by 1052 instead of 20 features
\cite{petritis2006improved}. The new descriptors encoded the identity
of each amino acid at each position in a sequence, as well as the
length and the hydrophobic moment of the peptide. The new neural
network was trained on more than 300,000 peptide sequences. To obtain
such a large number of peptides with known retention times, a strategy
to normalize elution times was developed. This allowed pooling of
peptides identified in more than 12,000 shotgun experiments. The
trained algorithm reported better performance for the 1303 peptides
on which it was evaluated, than its predecessor.

@@

\begin{figure}[p]
\centering 
\includegraphics[width=0.9\textwidth]{img/elude-ptm.pdf}
\caption{\label{fig:elude} {\bf {\sc Elude}'s workflow.}  {\sc Elude} encompasses two learning steps. In the first one the training peptides are used to build an in-house retention index. This index, together with additional information, is then employed to calculate a set of 60 numerical features for each peptide. These features are optimally combined during a second training procedure. The resulting retention model can be subsequently used to predict the retention time of other peptides of interest.}
\end{figure}

@@

The available technologies to infer peptides and proteins from mass
spectrometry data involve predicting the the potential manifestations
of the analyzed peptides, such as their fragmentation pattern, and
comparing those predicted properties to the observed data. Such
comparisons are regularly done for discovery proteomics by search
engines. However, an attractive way to increase the sensitivity of the
inference procedure is to also predict properties such as their
isoelectric point~\cite{perez2011silico,branca2014hirief} or of
particular interest for this review, chromatographic retention time
\cite{cerqueira2010mude}.

One way to determine a predictable separation techniques usefulness is
to estimate its ability to remove candidate peptides of any detected
feature, normally measured by a false positive ratio (FPR), also known
as the methods fall-out. The FPR is of course a function of the
selected thresholds, but if we select a treshold were say $\sim$95\%
of the known examples get correctly included, what FPR do we find at
such level?  Current retention time prediction algorithms have an FPR
of about $\frac{1}{5}$, meaning that it can remove about four out of
five candidate peptides for a given feature. This could be compared to
the FPR of about $\frac{1}{200}$ for predictions of high-resolution
iso-electric focusing experiments~\cite{branca2014hirief}. The
retention time prediction have to improve quite a bit to come to level
with the fall-out of isoelectric focusing, however, almost all mass
spectrometry-based proteomics use reversed phase chromatography, while
very few studies are using iso-electric focusing.

@@

%%  LocalWords:  hydrophobicity chromatography elute peptides physico
%%  LocalWords:  electrospray spectrometry analytical proteomics RPLC
%%  LocalWords:  posttranslational phosphorylation proteome analytes
%%  LocalWords:  peptide elutes amino tryptic isoelectric SSRCacls et
%%  LocalWords:  workflows trypsin hydrophobicities chromatographic
%%  LocalWords:  alkyl peptide's elution SSRCalc BioLCCC RTModel iRT
%%  LocalWords:  RTPredict Krokhin al QSRR Klammer SVMs SVM Moruz pre
%%  LocalWords:  dataset workflow outliers posttranslationally silico
%%  LocalWords:  phosphorylated artifactual phosphopeptides COmbined
%%  LocalWords:  FRActional DIagonal COFRADIC datasets Palmblad
%%  LocalWords:  Strittmatter unfragmented acetonitrile chromatogram
